Tracing the substrate translocation mechanism in P-glycoprotein

Version 1

Abstract

Abstract

P-glycoprotein (Pgp) is a prototypical ABC transporter of great biological and clinical significance that confers cancer multidrug resistance and mediates the bioavailability and pharmacokinetics of many drugs1–3. Decades of structural and biochemical studies have provided insights into how Pgp binds diverse compounds4–9, but how they are translocated through the membrane has remained elusive. Here, we covalently attached a cyclic substrate to discrete sites of Pgp and determined multiple complex structures in inward- and outward-facing states by cryoEM. In conjunction with molecular dynamics simulations, our structures trace the substrate passage across the membrane and identify conformational changes in transmembrane helix 1 (TM1) as regulators of substrate transport. In mid-transport conformations, TM1 breaks at glycine 72. Mutation of this residue significantly impairs drug transport of Pgp in vivo, corroborating the importance of its regulatory role. Importantly, our data suggest that the cyclic substrate can exit Pgp without the requirement of a wide-open outward-facing conformation, diverting from the common efflux model for Pgp and other ABC exporters. The substrate transport mechanism of Pgp revealed here pinpoints critical targets for future drug discovery studies of this medically relevant system.

Main Text

P-glycoprotein (Pgp, also known as MDR-1 and ABCB1) is a prominent member of the human ATP-binding cassette (ABC) efflux transporters that removes a large variety of chemically unrelated hydrophobic compounds from cells3, 9. Pgp is highly expressed in the liver, kidney, and intestines, where it limits absorption and enhances the excretion of cytotoxic compounds10–12. In the blood-brain, blood-testis, and blood-placenta barriers, Pgp protects and detoxifies those sanctuaries from xenobiotics13–15. Of central importance for many chemotherapeutic treatments, Pgp is a major determinant of drug bioavailability and pharmacokinetics and confers multidrug resistance in several diseases, most notably cancer16–18. Consequently, evasion, selective inhibition, and modulation of Pgp transport are important goals in drug development, but are hindered by a lack of detailed understanding of the drug transport mechanisms19.

For Pgp and ABC transporters in general, ATP binding, hydrolysis, and subsequent release of the cleaved products (inorganic phosphate and ADP) fuel large-scale conformational changes that ultimately result in translocation of substrates across the membrane bilayer20, 21. In inward-facing (IF) conformations, the nucleotide-binding domains (NBDs) are separated, and substrates can access a large binding cavity that is open to the lower membrane leaflet and cytoplasm. ATP-induced dimerization of the NBDs promotes large rearrangements of the transmembrane domains (TMDs) from an IF to an outward-facing (OF) conformation, from which the substrate is released22, 23. Pgp structures with bound substrates and inhibitors in IF conformations have revealed one or more overlapping binding sites6–8, 24, 25. Multimodal binding mechanisms for chemically related compounds have been proposed6, 9, including access of hydrophobic ligands from the lipid-protein interface to the binding sites of Pgp8, 24. However, structural data that elucidate how substrates are shuttled across the lipid bilayer to the extracellular space are unavailable at present. So far, only a single OF structure of mammalian Pgp has been described, but without detectable substrate density26. Most insights into the transport pathway arise from molecular dynamics simulations, which are also limited by the scarcity of available structural details27, 28. Consequently, the molecular mechanics that move the substrate through the transmembrane passage have remained elusive. To address this fundamental question and overcome the dynamic and transient nature of substrate translocation, we tethered a substrate molecule covalently to specific residues along the putative translocation pathway of Pgp and utilized cryoEM to capture Pgp structures with bound substrate during various stages of transport.

Covalent ligand design for Pgp labeling

For covalent labeling of Pgp, we synthesized a derivative of the previously published QZ-Ala tripeptide substrate24, by substituting one of the three Ala with Cys that is disulfide-linked to a 2,4-dinitrophenyl thiolate group (designated AAC-DNPT, See Methods, Extended Data Fig. 1). QZ-Ala is a strong ATPase stimulator and subject to Pgp transport as evidenced by its reduced cytotoxicity in cells overexpressing Pgp24. Additionally, we have measured Pgp-mediated transport of QZ-Ala in the MDCK-MDR1 monolayer permeability assay (see Methods). The efflux ratio (RE) of QZ-Ala was determined as 2.5. In the presence of the Pgp inhibitor cyclosporin A, RE was 0.9, indicating that the Pgp-specific transport of QZ-Ala was inhibited. Derivatizing QZ-Ala is simplified by its structural symmetry, which reduces the complexity of modifying this transport substrate for attaching to Pgp in different states. Furthermore, attachment of DNPT strongly activates disulfide exchange, allowing rapid and efficient crosslinking between the cyclic peptide and accessible cysteines. Free DNPT, released during crosslinking, is bright yellow. This provides a simple, visible readout that can be monitored with a UV-Vis spectrometer (Fig. 1a). Four single-Cys mutants of Pgp (Pgp335, Pgp978, Pgp971, and Pgp302) were generated for crosslinking, and mutations were chosen near the two previously reported QZ-Ala binding sites24. L335C in transmembrane helix 6 (TM6) and V978C in TM12 are situated at symmetric positions on opposing TMD halves, while L971C is located two helical turns above V978 (Fig. 1b). Residues L335C, V978C as well as I302C located in TM5 were previously shown to be labeled with verapamil29. Based on the appearance of the yellow DNPT byproduct, Pgp335 and Pgp978 reacted with AAC-DNPT within minutes, while Pgp971 and Pgp302 did not. Even after extended incubations only background absorbances were detected comparably to Cys-less Pgp (CL-Pgp) (Fig. 1c). Only after adding Mg2+ATP to fuel substrate translocation, Pgp971 could react with AAC-DNPT, but Pgp302 could not. Covalent attachment of the cyclic peptide AAC was further validated by high-resolution mass spectrometry of trypsin-digested peptide fragments (Extended Data Figs. 2 to 4). The presence of noncovalent transport substrates such as QZ-Ala or verapamil increase ATP hydrolysis in CL-Pgp and the single Cys mutants (Extended Data Fig. 5a). Similar to QZ-Ala, AAC-DNTP stimulated ATPase activity of CL-Pgp at low concentrations; with half-maximal enhancement of activity (EC50) seen at submicromolar concentrations, suggesting that it also acts as substrate (Extended Data Fig. 5b). At higher concentrations, the rate of ATP hydrolysis is significantly reduced in the mutants compared to CL-Pgp indicative of cross-linking AAC-DNPT with the single Cys mutants (Fig. 1d and Extended Data Fig. 5b). Importantly, Pgp335 and Pgp978 retain notably higher ATPase activity after cross-linking than the respective apo-Pgp mutants in the absence of substrate (Extended Data Fig. 5b and Fig. 1d), suggesting that tethered AAC still acts as a substrate and that Pgp335 and Pgp978 cycle between IF and OF conformations.

Covalent ligand design for Pgp transport studies.

(a) Design of AAC-DNPT, a derivative of the cyclic peptide QZ-Ala, for disulfide crosslinking with single-cysteine mutants of Pgp; yellow product DNPT used as readout. (b) Positions of L335 (TM6), V978 and L971 (TM12) in Pgp, residues targeted for cyclic peptide labeling. (c) Monitoring covalent reaction between AAC-DNPT and cysteine mutants (+/-Mg2+ATP) by UV-visible spectrometry; CL-Pgp and I302C were included as negative controls. (d) ATPase activities of L335C, V978C, and L971C before and after crosslinking with AAC-DNPT; CL-Pgp was used as a control. Data are averaged from triplicate measurements with standard error bars.

Pgp structures with bound substrates

To capture different conformations during the transport cycle of Pgp, we mutated the catalytic glutamate residues in the Walker-B motifs of each NBD (E552Q/E1197Q) in Pgp335 and Pgp978. When Mg2+ATP is present, these mutations arrest the enzyme in an ATP-occluded, NBD-dimerized conformation that resembles the pre-hydrolysis state26. For Pgp971, we prepared the covalent complex in the presence of Mg2+ ATP and stabilized the resulting OF complex with vanadate in a post-hydrolysis intermediate state30, 31 (Fig. 1c). CryoEM analysis of the respective covalent complexes in both OF and IF conformations provided eight high-resolution structures with zero, one, or two ligands bound. In addition, a control dataset with ATP, but without the substrate was collected for Pgp335 (Fig. 2, Extended Data Fig. 6 and 7, and Extended Data Table 1).

High-resolution structures of Pgp with bound ligands.

For all mutants, IF and OF structures were obtained, with ligands revealed in the transmembrane region (grey shadow). For Pgp335 three different OF conformations were detected in one data set (OF335-nolig, OF335-1lig, OF335-2lig). To display the binding sites, the upper halves of TM 2,10,11 and 12 were removed from the densities. For the IF structures, TM9 was also removed for clarity. TM1 is shown in brown, with the covalently bound ligand in red and the non-crosslinked in green. Significant conformational changes were observed among these structures from which a further zoom-in of TM1 (black box region) was shown in Fig. 4 and Extended Data Fig. 8.

The IF-structure of Pgp335, solved at 3.8 Å resolution, shows clear densities for two substrate molecules: one crosslinked and one not (IF335-2lig, Fig. 2 and Extended Data Fig. 7 and 8). The non-crosslinked molecule is bound in a position that resembles the one previously reported in a crystal structure24. It interacts closely with the covalently-linked substrate, which is situated further up the membrane. Both ligand molecules are oriented along the vertical axis of the protein and positioned side-by-side through hydrophobic interactions. The uncleaved DNPT group of the non-crosslinked substrate is evident in the cryoEM density map. Two substrate molecules are also bound in the IF structure of Pgp978 (IF978-2lig); the non-covalently bound molecule is clearly visible at a similar position as in IF335-2lig, while the crosslinked substrate is located on the opposite side of the binding cavity near TM12 (Extended Data Fig. 8). Corroborating our labeling data (Fig. 1c), in the IF structure of Pgp971 (IF971-1lig) at 4.3 Å resolution, only the non-crosslinked substrate is bound at the approximate location seen for non-crosslinked substrate in IF335-2lig and IF978-2lig (Fig. 2, Extended Data Figs. 8 and 9e).

Multi-model cryoEM analyses of ATP-bound Pgp335 revealed three different OF conformations within a single dataset: ligand-free (42%), single-ligand-bound (34%), and double-ligand-bound (26%) (OF335-nolig, OF335-1lig and OF335-2lig), at 2.6 Å, 2.6 Å, and 3.1 Å resolution, respectively (Extended Data Fig. 10, Fig. 2). For Pgp978 and Pgp971, only OF conformations with a single bound ligand (OF978-1lig and OF971-1lig, Fig. 2) were detected at 2.9 Å and 3 Å resolution, respectively.

Substrate translocation pathway

In the OF structures with a single bound substrate, the central transmembrane helices TM1, TM6, TM7 and TM12 bulge outwards to accommodate the substrate in a small tunnel (Extended Data Fig. 11). Comparing the ligand positions between the IF and OF structures, in IF335-2lig to OF335-1lig, and IF978-2lig to OF978-1lig, the covalently attached ligand pivots around the respective Cys residue by almost 180°, moving it further up the translocation tunnel (Extended Data Figs. 8 and 9a,b, and Supplementary Videos 1 and 2). Our data do not reveal whether rotation of the ligand during the IF-OF conformational change is required for transport or if this reflects the fact that the ligand is covalently pinned to the crosslinking residue. However, the observation of ligand rotation illustrates how much space is available in the binding cavity during this conformational transition. The ligand position and orientation in OF335-1lig and OF978-1lig overall coincide well, with the latter shifting up slightly (Extended Data Fig. 9d). This finding indicates that the export tunnel is inherent to substrate translocation regardless of the initial binding position in the IF conformation. As such, our data suggest a single common pathway for substrate export in Pgp, contrary to the dual pseudosymmetric pathways suggested earlier28, 32.

OF971-1lig captures a later-stage transport intermediate, in which the compound resides approximately 7 Å further up the translocation tunnel and is coordinated by highly conserved residues, including M74 and F78 (Figs. 2 and 3b, Extended Data Figs. 8 and 12). As no crosslinked substrate was detected in the corresponding IF structure, the substrate was likely transported to residue L971C before the crosslinking reaction could take place. Substrate entry from the extracellular side is less likely given the small opening of the substrate translocation pathway in all OF Pgp structures with or without bound substrate obtained here and previously26. In all substrate bound OF conformations methionines and phenylalanines at different positions seem to be important for ligand coordination in all of the Pgp structures (Extended Data Fig. 12). As expected, the substrate is held in the binding pocket mainly by the hydrophobic interactions throughout the different conformations (Extended Data Figs. 8 and 12). In the OF335-nolig structure, as well as in the virtually indistinguishable control, the translocation tunnel is collapsed, which shields the substrate binding pocket from the extracellular milieu and prevents re-entry from this side26.

Functional characterization of G72A mutant.

(a) TM1 breaks at the G72 position in the mid-transport states of Pgp. Shown is a close-up view of TM1 (brown) with side chain densities, the G72 position (green arrow), and the adjacently bound cyclic peptide AAC (red sticks) in the OF335-1lig structure. (b) Sequence alignment of TM1 in vertebrate Pgps including the highly conserved G72 residue (green arrow). (c) Level of expression of the G72A mutant relative to the WT- and CL-Pgp in S. cerevisiae cultures derived from the western blot analysis. (d) In vivo activity of the G72A mutant. The growth resistance of S. cerevisiae cultures transformed with either vector control plasmid, WT-Pgp, CL-Pgp, or the G72A mutant against several fungicides (doxorubicin, FK506, and valinomycin) relative to the YPD medium only (black bar) were analyzed. Compared to WT-Pgp and CL-Pgp, the G72A mutant severely compromised growth resistance against the tested drugs (p-values of <0.001 are indicated by ***). (e-g) ATPase activity of the purified G72A mutant. While basal ATP hydrolysis rates (in the absence or at low concentrations of drug) were indistinguishable between WT and G72A mutant proteins, stimulation of the ATPase activities by 1-3 µM FK506, 1-3 µM valinomycin, or 30-130 µM verapamil were severely impaired (p-values of <0.001 are indicated by ***).

Regulatory role of TM1

The series of IF and OF Pgp structures reveal a cascade of conformational changes in TM1, which displays significant plasticity throughout the different transport stages (Fig. 2, Extended Data Fig. 8, and Supplementary Videos 1, 2 and 3). In IF conformations, TM1 is a long, straight helix (Fig. 2 and Extended Data Fig. 8). However, in OF335-1lig, OF978-1lig, and OF971-1lig, TM1 swings out at A63, dilating the transmembrane passage and providing sufficient space for accommodating the bound ligand. At the same time, a pronounced kink at G72 on TM1 destabilizes the helix and the extracellular loop of Pgp between TM1 and TM2 and leads to a partial unwinding. This conformational deformation shields the intramembranous tunnel from the extracellular environment, with TM1 acting as a lid, almost parallel to the membrane (Fig. 2 and 4a, Extended Data Fig. 8). Strikingly, along with the upward movement of the ligand (from L335C to V978C to L971C), the TM1 loop is progressively lifted and restabilized. Ultimately, as shown previously26 and by our OF335-nolig structure, the helix is straightened upon the release of the ligand (Fig. 4a, Extended Data Fig. 8). Interestingly, the extracellular gate is sealed in all OF structures, which is especially surprising for OF971-1lig where the ligand is near the tunnel exit (Fig. Extended Data Fig. 11).

Tracing the substrate movement in Pgp.

Structures of OF Pgp from cryoEM (a) aligned with four snapshots from MD simulations (b) of substrate transport and release. TM1 is rendered in brown, with A63 and G72 in green, and M68, M74, and F78 in plum. The cyclic peptide is shown as red sticks. Significant conformational changes were observed for TM1 along with the upward movement and release of the substrate from OF Pgp.

Using OF978-1lig as initial configuration, we performed molecular dynamics (MD) simulations of Pgp embedded in a lipid bilayer to validate the proposed substrate escape pathway. Accelerated by temperature (T=400 K), we observed movement and release of the non-covalently bound substrate to the extracellular side moderated by interactions between the ligand and hydrophobic residues in the upper part of TM1 (Supplementary Video 3). The intermediate steps coincided well with the conformations observed in cryoEM (Fig. 4, Extended Data Figs. 13 and 14). Without any bias imposed, the substrate progressed to a position closely matching OF971-1lig (Fig. 2, 4b). Then, while TM1 gradually refolded from the intracellular side, the substrate wiggled out to form transient interactions with the still unstructured extracellular part of TM1 (Fig. 4). Importantly, the substrate escaped without Pgp experiencing a wide opening of the extracellular region. After the substrate left the exit tunnel, TM1 continued to refold to attain a straight helix configuration like OF335-nolig at the end of the simulation (Fig. 4). Overall, with exception of TM1, motions in Pgp to accommodate the moving substrate were mostly limited to small dilations of the transmembrane helices, supporting our observations from cryoEM.

Our cryoEM structures and MD simulations revealed an important regulatory role of TM1 during substrate transport, emphasizing the helix-breaking G72. Therefore, we mutated this residue, which is highly conserved in mammalian Pgp (Fig. 3), to a helix-stabilizing alanine33 and tested the mutant’s ability to export structurally diverse fungicides and to convey drug resistance in vivo. Compared to the wild-type and CL-Pgp, the G72A mutant expressed normally in Saccharomyces cerevisiae (Fig. 3c), but showed significantly reduced growth resistance to the test drugs including doxorubicin, FK506, and valinomycin (Fig. 3d). As drug transport was severely impaired, we further investigated the mutant’s ATPase activity. In presence of transport substrates such as FK506, valinomycin, or verapamil the ATPase activity of WT Pgp was significantly stimulated; however, the G72A mutant displayed only low levels of basal ATPase activity (in the absence of drug) and the response to drug stimulation was greatly diminished over a range of concentrations tested (Fig. 3e-g). To reveal potential structural consequences responsible for the loss of function, we performed cryoEM on the Pgp335 G72A mutant after crosslinking the single cysteine L335C with AAC-DNTP and stabilizing the OF complex with Mg2+ ATP and vanadate in a post-hydrolysis intermediate state. After multi-model classification, two conformations emerged. The first conformation, resolved at 3.0 Å, shows no substrate bound in analogy to OF335-nolig. The second dominant conformation is, however, much more heterogenous, resulting in a reduced and anisotropic resolution of approximately 4.6 Å (Extended Data Fig. 15). Here, the NBDs are well resolved but several transmembrane helices are missing, including TM1 in which residue G72A is located. The structural disorder, observed in the transmembrane region, is likely a result of the tethered ligand, causing local distortion of the surrounding α-helices as TM1 may not be able to kink at G72 in the mutant, which could explain its impaired transport function. Collectively, the significant differences in drug resistance profiles and drug-stimulated ATP hydrolysis rates between the G72A mutant and WT-Pgp along with our structural findings substantiate and highlight the key role of TM1 for the regulation of substrate transport in Pgp.

Discussion

Our data suggest a mechanism for how Pgp mediates substrate transport. Transport commences in the IF conformation with binding of the substrate. Upon ATP-induced NBD-dimerization, the transporter converges into an OF conformation in which the intracellular gate is closed. The presence of the substrate in the OF conformation of Pgp dislocates TM1 at A63 and introduces a sharp kink at G72, forcing the helix to unwind partially between TM1 and TM2 (Fig. 4 and Extended Data Fig. 8). This enlarges the extracellular loop which shields the transmembrane tunnel from the extracellular environment and guides the substrate to move up the translocation tunnel, allowing TM1 to gradually restabilize. During the last stages of transport (OF971-1lig to OF335-nolig), the TM1-TM2 regulatory loop straightens, and the substrate-interacting hydrophobic residues (M74 and F78) rotate and pivot away from the tunnel, which facilitates the escape of the substrate from the transporter. Although OF971-1lig captured a post-hydrolysis state, we cannot rule out the possibility of ligand release prior to ATP hydrolysis, when it is not covalently bound. Notwithstanding, the proposed TM1 kinking and straightening mechanism of substrate expulsion would obviate the requirement for the Y-shaped, wide-open OF conformation reported for multiple bacterial ABC transporters34, 35, suggesting a substrate-dependent transport mechanism in Pgp, or else a divergence of mechanisms among ABC transporters.

For the OF335-2lig structure, observed in the same data set as OF335-1lig and OF335-nolig, we hypothesize that the hydrophobic nature of the substrate promotes the binding of two substrate molecules to each other. MD simulations with this structure as a starting conformation reveal no separation of the two molecules, suggesting that they might be treated as one bigger substrate in the transport event. With the binding positions differing between OF335-1lig and OF335-2lig (Extended Data Fig. 9c, 12 and 16), neither the cryoEM structure nor the simulations allow conclusions about a possible transport mechanism of two ligands at this point.

How ABC transporters move substrates through the lipid bilayer has been a long-standing question36. The scarcity of high-resolution structures of substrate-bound OF conformations37, especially in the highly transient intermediate states, has severely limited our understanding of this vital process. Here, we were able to determine eight high-resolution structures of the archetypical ABC transporter Pgp, revealing the underlying molecular mechanics of substrate translocation. In general, the large number and diversity of ABC transporters, and the even wider variety of their substrates, make it unlikely that a universal substrate translocation mechanism will emerge. Our strategy of stalling the otherwise highly transient intermediates in this process through crosslinking and subsequent high-resolution cryoEM analysis could serve as a blueprint to understand the transport mechanisms of lipid transporters as well as other members of the transporter family.

Supporting information

eLife Assessment

P-glycoprotein (Pgp) is a major ABC-transporter that couples ATP hydrolysis with the export of xenobiotics and drugs, influencing their pharmacokinetics. The authors provide convincing cryo-EM Pgp structures of drug complexes in previously unforeseen outward-facing conformations. These structures reveal important mechanistic insights that could be of broader impact, but because these transient-like states were captured by cysteine cross-linking the substrate to Pgp, support for the conclusions is incomplete, and further validation is required.

Reviewer #1 (Public Review):

Summary:

Here the authors have tethered a Pgp substrate to strategically place cysteine residues in the protein. Notably, the cysteine-linked substrate (ANC-DNPT)- stimulates ATP hydrolyse and so is able to undergo IF to OF transitions. The authors then determined cryo-EM structures of these complexes and MD simulations of bound states. By capturing unforeseen OF conformations with substate they propose that TM1 undergoes local conformational changes that are sufficient to translocate substrates, rather than large bundle movements.

Strengths:

This paper provides the first substrate (ANC-DNPT)- bound conformations of PgP and a new mechanistic model of how substrates are translocated.

Weaknesses:

Although the cross-links stimulate ATP hydrolysis, further controls are needed to convince me that the TM1 conformations observed in the structures are physiologically relevant, since they have been trapped by "large" substrates covalently-tethered by cross-links.

Reviewer #2 (Public Review):

Summary:

The manuscript by Gewering and coworkers is an elegant mechanistic investigation of the mammalian multidrug transporter Pgp. I will not elaborate on the significance of this protein except to point out its clinical involvement in cancer resistance to chemotherapy.

Strengths:

The strengths of the investigation are partly in the combination of sophisticated chemical synthesis, state-of-the-art cryoEM in a well-established biochemical context. What is more exciting is the tackling of a long-standing question in the field: namely how do drugs make their way through the structure to be exported across the membrane? Unfortunately, the field has been stuck in hand waving model based on structures that in the outward-facing conformations are devoid of substrates. The work challenges the dogma that emerged from this hand-waving model and presents an alternative model that appears to be supported by the data.

Weaknesses:

There is much to like about the experimental work here but I am less sanguine on the interpretation. The main idea is to covalently link via disulfide bonds a model tripeptide substrate under different conditions that mimic transport and then image the resulting conformations. The choice of the Pgp cysteine mutants here is critical but also poses questions regarding the interpretation. What seems to be missing, or not reported, is a series of control experiments for further cysteine mutations.

Reviewer #3 (Public Review):

Summary:

The authors employed a new strategy, covalent substrate-labeling, to address the open issue of the substrate transport mechanism by single particle cryogenic-EM. A cyclic peptide (QZ-Ala), which was already used in the past as a substrate for structural purposes, was modified and covalently attached to ABCB1 at strategic positions in the transmembrane domain via Cys-specific coupling chemistry. Overall, four mutations (two per TMD) were generated and functionally analyzed. These residues are in proximity to the QZ-Ala binding site and are labeled by verapamil. Interestingly, two mutants could only be labeled if ATP and Mg2+ were present.

Strengths:

Three of the four mutants were structurally analyzed by single particle cryo-EM with structures in both, the inward- and outward-facing conformation and overall resolutions ranging from 2.6 to 4.3 Å. Applying multi-model analyses allowed for the extraction of additional structures from one data set. These structures formed the basis for a detailed analysis of the substrate translocation pathway. This enabled the researcher to compare the IF and OF states of the same mutant and same substrate, which is pivotal for their conclusions. The mutations 335/978 trap the substrate at different points of the translocation pathway, while 971 located two helical turns away from the first set, trapped the system at a later stage. The described strategy revealed a cascade of conformational changes during substrate transport which focus on TMH1, which is straight in the IF state, but swings out in the OF state. Pivotal for such a change is G72. This residue was mutated to Ala and also structurally analyzed. These structures were supported by MD simulations and functional data. Thus, the new prosed kinking and straightening mechanisms are different from the so far accepted wide-open OF state observed in bacterial transporters. This clearly suggests a different mechanism for ABCB1.

Weaknesses:

I have a couple of minor issues that I have listed in the section recommendations for authors Overall, the manuscript is very well written, sheds new light on the molecular mechanism of substrate translocation by ABCB1, and might even provide a new starting point for inhibitor design. I know that it is unusual, but I like the manuscript in its current version and recommend acceptance in its current form.

Author Response

The following is the authors’ response to the original reviews.

Reviewer 1 (Public Review):

Weakness: Although the cross-links stimulate ATP hydrolysis, further controls are needed to convince me that the TM1 conformations observed in the structures are physiologically relevant, since they have been trapped by "large" substrates covalently-tethered by crosslinks.

Our response: Reviewer 1 raised concerns about the relatively large size of our covalently attached AAC substrate that would potentially distort TM1 in Pgp. We would like to clarify that AAC has a molecular weight of 462 Da, which, in comparison to many known Pgp substrates ranging from 250 to over 1,000 Da, is not a large compound. For instance, the few other Pgp substrates mentioned in our manuscript all have a comparable or larger size: verapamil, 455 Da; doxorubicin, 544 Da; FK506, 804 Da; valinomycin, 1,111 Da; cyclosporin A, 1,203 Da.

Furthermore, AAC was strategically attached to a site distant from TM1 in the inwardfacing Pgp conformation. After it was exported to the outward-facing state, several TM helices accommodate the compound. The observation that only TM1 exhibited significant conformational changes suggests its potential role in the transport mechanism. This hypothesis is supported by our findings, where a conservative substitution (G72A) in TM1 resulted in a dramatic loss of transport function for various drug substrates and impaired verapamil-stimulated ATPase activity.

Reviewer 1 (Recommendations for the Authors):

I understand the need for an unconventional approach to understanding the translocation pathway. What would help to support this model is to cross-link a much smaller substrate, as the one used is quite large and could potentially distort TM1 in the outward-state when cross-linked.

Our response: We thank the reviewer for this recommendation, and we have outlined plans for future experiments involving other substrates, including smaller ones, to further investigate our proposed model. However, it is important to acknowledge that conducting these studies will require a significant amount of effort and resources, which we believe extend beyond the scope of our current manuscript.

In unbiased MD simulations starting from the IF state are there any simulations where the substrate follows the same path as proposed here?

Our response: All our MD simulations were performed in the outward-facing state to focus on potential substrate release pathways. Starting MD simulations from the inwardfacing state would introduce complexities in capturing the necessary domain motions and nucleotide binding and hydrolysis required for substrate translocations. Therefore, we opted not to perform MD studies starting from the inward-facing state.

Reviewer 2 (Public Review):

Weakness: There is much to like about the experimental work here but I am less sanguine on the interpretation. The main idea is to covalently link via disulfide bonds a model tripeptide substrate under different conditions that mimic transport and then image the resulting conformations. The choice of the Pgp cysteine mutants here is critical but also poses questions regarding the interpretation. What seems to be missing, or not reported, is a series of control experiments for further cysteine mutations.

Our response: Reviewer 2 raised concerns about the interpretation of our results and suggested the need for additional mutant designs to validate our proposed TM1 mechanism. Firstly, we believe that the observed TM1 conformational changes are valid in our cryoEM structures, despite the use of different conditions and several mutants to capture Pgp in the outward-facing state.

Regarding the G72A mutant, we consider it conclusive that this single point mutation in the TM1 has a profound effect. Importantly, the G72A mutant was readily expressed and purifiable as a stable protein. We were able to resolve a high-resolution structure of the G72A mutant (without the substrate), confirming that the protein is not generally destabilized but properly folded.

Above all, we appreciate the Reviewer’s suggestion to explore additional mutations and intend to do so in future studies.

Reviewer 2 (Recommendations for the Authors):

I am sold on the results regarding TM1 conformational changes as they are evident in the cryoEM structures. However, the set of states compared between mutants are not biochemically equivalent: for 335 and 978 they used an ATP-impaired Pgp whereas for 971 they used what appears to be WT, and the conformation was imaged presumably subsequent to ATP hydrolysis and Vanadate trapping. This is significant if the authors were unable to trap the OF in the impaired mutant background and should be highlighted. I have to believe that they tried that condition but I could be wrong.

Our response: We acknowledge the point made by the Reviewer about the biochemical equivalence of mutant states and the potential significance of using an ATP-impaired mutant for trapping the outward-facing conformation of 971. We have not yet attempted to use the ATPase-deficient 971C mutant for crosslinking and intend to address this question in future studies.

In our current approach, we used the ATPase-active 971C for two specific reasons:

1. Our biochemistry data, as shown in Fig 1C, indicates that 971C only crosslinks in the presence of ATP hydrolysis conditions. Vanadate trapping was employed to stabilize the outward-facing conformation.

2. Based on our experience, we have observed that the conformations of ATP-bound (mutant) and vanadate-trapped states of an ABC transporter are structurally equivalent at this resolution level of our study (see ref. 21: Hoffmann et al. NATURE 2019).

The authors propose a new model for substrate translocation. It is based on three mutants and a number of structures. If the authors were not challenging the current dogma I would not have written the next comment. Considering the impact of the findings, I would have designed a couple more cysteine mutants based on their model. For instance, this pathway has a number of stabilizing interactions, can't they make a mutant that preserves conformational switching but eliminates substrate translocation? I like the G97A mutant result but I am worried that the effect could just be a general destabilization or misfolding as part of the cryoEM particles seem to suggest. The authors advance one interpretation of the disorder observed in this mutant but it could easily be my interpretation.

Our response: We thank the reviewer for the suggestion to design additional mutants to further validate our proposed model for substrate translocation. We agree that this would be highly valuable, considering the potential impact of our findings. However, given the time-intensive nature of our approach, we believe that presenting these additional designs in a future study is a reasonable course of action.

Regarding the G72A mutation, we believe that our current data fully supports our model and the role of TM1 in regulating the Pgp activity. Importantly, we would like to emphasize that the G72A mutant was readily expressed and purifiable as a stable protein. Additionally, our cryoEM structural determination of the G72A mutant at high resolution confirmed that the protein is not generally destabilized but properly folded.

There are a couple of troubling methodological questions that I want the authors to address or clarify:

1. In the methods they report that the final sample for cryoEM was prepared on a SEC devoid of detergent. It is obvious that the sample was folded but I was wondering why the detergent was removed? Was that critical for observing these structures with multiple ligands? Did they observe any lipids in their cryoEM?

Our response: We avoid detergent in the buffer on final SEC purification. This step is to remove free detergent from the background which helps during cryoEM imaging. Of course, this cannot be done with every detergent but due to the very low CMC of LMNG it is possible. By now, we have verified this method for several other transporters with the same success. While this procedure helps us to obtain better images it is not necessary to obtain specific conformations or ligand bound states, nor does it affect these states or conformations.

In our cryoEM structures , we did observe multiple cholesterol hemisuccinate (CHS) molecules on the outer transmembrane surface of Pgp.

1. Can the authors comment on why labeling was carried out in the presence of ATP? Does it matter if the substrate was added prior to ATP and incubated for a few minutes?

Our response: For every dataset, we first added the substrate to be cross-linked and afterwards added the ATP. In the cases of 335C and 978C, labeling was successful before ATP was added, as evidenced by the inward-facing structures with cross-linked substrate. However, for 971C, cross-linking only occurred after the addition of ATP. We interpret this data to suggest that the 971 site is inaccessible to the substrate in the inward-facing state, and cross-linking can only occur after the transporter transitions to outward-facing state. This is in line with our inward-facing structure which does not show a cross-linked substrate, and our biochemical data shown in Fig 1C, where 971C only crosslinked in the presence of ATP.

1. I am not an expert on MD simulations and I understand that carrying out simulations at higher temperatures used to be a trick to accelerate the process. Is this still necessary? Why didn't the author use approaches such as WESTPA?

Our response: Most so-called enhanced sampling methods, including WESTPA, explicitly define a reaction coordinate for the process of interest, usually based on intuition or prior studies. If this coordinate is chosen poorly, enhanced sampling usually fails, either because the sampling becomes inefficient or because the sampling biases the transition pathway (or both). Lacking reliable intuition or prior knowledge on which motions would result in substrate release, we chose temperature to speed up the process. High temperature largely avoids the introduction of an any bias through the definition of a progress coordinate. By contrast, the weighted ensemble method underlying WESTPA is a great method to simulate unbiased dynamics of a process with a known progress coordinate, but unfortunately requires to choose a progress coordinate prior to the simulation and will then mostly sample the process along this progress coordinate, because this is the only direction in which sampling is improved. High temperature MD on the other hand accelerates all processes in the system under study. Indeed, we have now confirmed that the pathway found at high temperature is also feasible at near-ambient conditions.

In new simulations, we have now observed a similar release pathway at T=330 K. As the only difference, the substrate has not fully dissociated from the protein after 2.5 us, with weak interactions persisting at the top part of TM1 from the extracellular side. Importantly, this is a configuration observed also in higher temperature simulations but with much shorter lifetime.

In response, we now included these new findings and a new Extended Data Fig. 15 in the revised manuscript.

1. One way to show that the two substrates binding mode is biochemically relevant is to measure Vmax at different substrate concentrations. One would expect a cooperative transition if that interaction is mechanistically important.

Reviewer 3 (Public Review):

We thank Reviewer 3 for recommending the acceptance of our manuscript as is.

Reviewer 3 (Recommendations for the Authors):

Page 4, last line: Pgp302 should be Pgp1302. In addition, I can only encourage the authors to add an additional table to the manuscript. Here, the mutation, the obtained structure(s), IF or OF, the resolution, and the main message should be summarized.

Our response: Following the reviewer’s suggestion, we have added Extended Data Table 2 summarizing the Pgp mutants and respective structural data in the revised manuscript.

We verified that Pgp302 is the correct term on Page 4, last line.

Pg. 5, section 'Covalent ligand design for Pgp labeling', it is mentioned that even in the presence of Mg2+ATP, Pgp302 could not react with AAC-DNPT. Maybe it would be worthwhile to add the data either in Supplementary Information or state 'data not shown'.

Our response: We stated ‘data not shown’ in the text.

Pg. 47, last line : A space is missing between M68, and M74.

Our response: Space was added.

Pg. 7, line 2: The authors mention that a single dataset of ATP-bound Pgp335 revealed three different OF conformations: ligand-free, single-ligand-bound, and double-ligandbound. However, the percentage fraction of each dataset sums up to be more than 100%. Would request the authors to recalculate the fraction size of each conformation.

Our response: We have corrected the error in our calculation, based on the particle distribution in our dataset (OF335-nolig: 1,437,110 particles, 40.4%; OF335-1lig: 1,184,253 particles, 33.3%; and OF335-2lig: 939,924 particles, 26.4%).

Pg 53, Figure legend of Extended Data Fig. 11: Please include the color coding for the helix TM1 and also the residues colored plum.

Our response: We added the color coding for TM1 and other residues in the figure legend.

Pg. 8, line 3: While referring to the structure of OF971-1lig, the authors nicely point towards the conserved residues M74 and F78 which coordinate the ligand. However, in Fig. 3b, residues M74 and F78 should also be indicated.

Our response: We updated Fig. 3b by adding arrows pointing towards the residues M74 and F78.

Pg. 54, Extended data Fig. 12: The authors should adopt a single writing style. In some places, Pgp is referred to as P-gp while in others as Pgp.

Our response: We updated the protein labels in Extended Data Fig. 12.

Pg. 54, Extended data Fig. 12: The authors should clearly mention which OF335 structure (1st panel) was used for visualizing the interactions.

Our response: To clarify, we added the following sentences in the figure legend: “Pgp335 OF in the top panel refers to OF335-1lig. In the bottom panel describing OF335-2lig, the left and right diagrams refer to the binding positions of non-covalent and covalent ligand, respectively”.

Pg. 18, section 'synthesis of dipeptide 8': In the text it is mentioned that for the synthesis of thiazole acid 6, compound 3 was dissolved in a mixture of THF/MeOH/H2O (3:1:1), while in the corresponding figure (Extended Data Fig. 1), the ratio is stated as 5:1:2.

Our response: 3:1:1 ratio is correct. We made the correction in Extended Data Fig. 1.

Pg. 19, section 'synthesis of linear tripeptide 10': Same as above for compounds 10 and 4, respectively.

Our response: We corrected the conditions in the Extended Data Fig. 1 accordingly.

Pg. 20, section 'Synthesis of cyclic peptide 11': There seems to be a discrepancy in the synthesis protocol between the text and the extended figure 1, especially regarding the use of THF/MeOH/H20, followed by NaOH and TFA or only NaOH and TFA.

Our response: we further clarified the conditions of using NaOH in THF/MeOH/H2O (3:1:1) and TFA in DCM in the text for synthesis and Extend Data Fig. 1.

Pg. 40, Extended Data Fig. 1: In the bottom last panel showing the synthesis of peptide 11, the authors have missed showing peptide 10 as the starting material for the reaction.

Our response: Label for the peptide 10 was added following the suggestion.

Pg. 26, third last line: 'o' is missing from the last word cry'o'

Our response: We corrected the typo.

Pg. 63 and 64, Extended Data Table 1: The Cryo-EM data collection, refinement, and validation statistics for OF971-1lig, IF971-1lig, OF978-1lig, and IF978-2lig are mentioned twice in the table.

Our response: This was now corrected in the revision.

Version 2

Abstract

Abstract

P-glycoprotein (Pgp) is a prototypical ABC transporter of great biological and clinical significance that confers cancer multidrug resistance and mediates the bioavailability and pharmacokinetics of many drugs1–3. Decades of structural and biochemical studies have provided insights into how Pgp binds diverse compounds4–9, but how they are translocated through the membrane has remained elusive. Here, we covalently attached a cyclic substrate to discrete sites of Pgp and determined multiple complex structures in inward- and outward-facing states by cryoEM. In conjunction with molecular dynamics simulations, our structures trace the substrate passage across the membrane and identify conformational changes in transmembrane helix 1 (TM1) as regulators of substrate transport. In mid-transport conformations, TM1 breaks at glycine 72. Mutation of this residue significantly impairs drug transport of Pgp in vivo, corroborating the importance of its regulatory role. Importantly, our data suggest that the cyclic substrate can exit Pgp without the requirement of a wide-open outward-facing conformation, diverting from the common efflux model for Pgp and other ABC exporters. The substrate transport mechanism of Pgp revealed here pinpoints critical targets for future drug discovery studies of this medically relevant system.

Main Text

P-glycoprotein (Pgp, also known as MDR-1 and ABCB1) is a prominent member of the human ATP-binding cassette (ABC) efflux transporters that removes a large variety of chemically unrelated hydrophobic compounds from cells3,9. Pgp is highly expressed in the liver, kidney, and intestines, where it limits absorption and enhances the excretion of cytotoxic compounds10–12. In the blood-brain, blood-testis, and blood-placenta barriers, Pgp protects and detoxifies those sanctuaries from xenobiotics13–15. Of central importance for many chemotherapeutic treatments, Pgp is a major determinant of drug bioavailability and pharmacokinetics and confers multidrug resistance in several diseases, most notably cancer16–18. Consequently, evasion, selective inhibition, and modulation of Pgp transport are important goals in drug development, but are hindered by a lack of detailed understanding of the drug transport mechanisms19.

For Pgp and ABC transporters in general, ATP binding, hydrolysis, and subsequent release of the cleaved products (inorganic phosphate and ADP) fuel large-scale conformational changes that ultimately result in translocation of substrates across the membrane bilayer20,21. In inward-facing (IF) conformations, the nucleotide-binding domains (NBDs) are separated, and substrates can access a large binding cavity that is open to the lower membrane leaflet and cytoplasm. ATP-induced dimerization of the NBDs promotes large rearrangements of the transmembrane domains (TMDs) from an IF to an outward-facing (OF) conformation, from which the substrate is released22,23. Pgp structures with bound substrates and inhibitors in IF conformations have revealed one or more overlapping binding sites6–8,24,25. Multimodal binding mechanisms for chemically related compounds have been proposed6,9, including access of hydrophobic ligands from the lipid-protein interface to the binding sites of Pgp8,24. However, structural data that elucidate how substrates are shuttled across the lipid bilayer to the extracellular space are unavailable at present.

So far, only a single OF structure of mammalian Pgp has been described, but without detectable substrate density26. Most insights into the transport pathway arise from molecular dynamics simulations, which are also limited by the scarcity of available structural details27,28. Consequently, the molecular mechanics that move the substrate through the transmembrane passage have remained elusive. To address this fundamental question and overcome the dynamic and transient nature of substrate translocation, we tethered a substrate molecule covalently to specific residues along the putative translocation pathway of Pgp and utilized cryoEM to capture Pgp structures with bound substrate during various stages of transport.

Covalent ligand design for Pgp labeling

For covalent labeling of Pgp, we synthesized a derivative of the previously published QZ-Ala tripeptide substrate24, by substituting one of the three Ala with Cys that is disulfide-linked to a 2,4-dinitrophenyl thiolate group (designated AAC-DNPT, See Methods, Extended Data Fig. 1). QZ-Ala is a strong ATPase stimulator and subject to Pgp transport as evidenced by its reduced cytotoxicity in cells overexpressing Pgp24. Additionally, we have measured Pgp-mediated transport of QZ-Ala in the MDCK-MDR1 monolayer permeability assay (see Methods). The efflux ratio (RE) of QZ-Ala was determined as 2.5. In the presence of the Pgp inhibitor cyclosporin A, RE was 0.9, indicating that the Pgp-specific transport of QZ-Ala was inhibited. Derivatizing QZ-Ala is simplified by its structural symmetry, which reduces the complexity of modifying this transport substrate for attaching to Pgp in different states. Furthermore, attachment of DNPT strongly activates disulfide exchange, allowing rapid and efficient crosslinking between the cyclic peptide and accessible cysteines. Free DNPT, released during crosslinking, is bright yellow. This provides a simple, visible readout that can be monitored with a UV-Vis spectrometer (Fig. 1a). Four single-Cys mutants of Pgp (Pgp335, Pgp978, Pgp971, and Pgp302) were generated for crosslinking, and mutations were chosen near the two previously reported QZ-Ala binding sites24. L335C in transmembrane helix 6 (TM6) and V978C in TM12 are situated at symmetric positions on opposing TMD halves, while L971C is located two helical turns above V978 (Fig. 1b). Residues L335C, V978C as well as I302C located in TM5 were previously shown to be labeled with verapamil29. Based on the appearance of the yellow DNPT byproduct, Pgp335 and Pgp978 reacted with AAC-DNPT within minutes, while Pgp971 and Pgp302 did not. Even after extended incubations only background absorbances were detected comparably to Cys-less Pgp (CL-Pgp) (Fig. 1c). Only after adding Mg2+ATP to fuel substrate translocation, Pgp971 could react with AAC-DNPT, but Pgp302 could not (data not shown). Covalent attachment of the cyclic peptide AAC was further validated by high-resolution mass spectrometry of trypsin-digested peptide fragments (Extended Data Figs. 2 to 4). The presence of noncovalent transport substrates such as QZ-Ala or verapamil increase ATP hydrolysis in CL-Pgp and the single Cys mutants (Extended Data Fig. 5a). Similar to QZ-Ala, AAC-DNTP stimulated ATPase activity of CL-Pgp at low concentrations; with half-maximal enhancement of activity (EC50) seen at submicromolar concentrations, suggesting that it also acts as substrate (Extended Data Fig. 5b). At higher concentrations, the rate of ATP hydrolysis is significantly reduced in the mutants compared to CL-Pgp indicative of cross-linking AAC-DNPT with the single Cys mutants (Fig. 1d and Extended Data Fig. 5b). Importantly, Pgp335 and Pgp978 retain notably higher ATPase activity after cross-linking than the respective apo-Pgp mutants in the absence of substrate (Extended Data Fig. 5b and Fig. 1d), suggesting that tethered AAC still acts as a substrate and that Pgp335 and Pgp978 cycle between IF and OF conformations.

Covalent ligand design for Pgp transport studies.

(a) Design of AAC-DNPT, a derivative of the cyclic peptide QZ-Ala, for disulfide crosslinking with single-cysteine mutants of Pgp; yellow product DNPT used as readout. (b) Positions of L335 (TM6), V978 and L971 (TM12) in Pgp, residues targeted for cyclic peptide labeling. (c) Monitoring covalent reaction between AAC-DNPT and cysteine mutants (+/- Mg2+ATP) by UV-visible spectrometry; CL-Pgp and I302C were included as negative controls. (d) ATPase activities of L335C, V978C, and L971C before and after crosslinking with AAC-DNPT; CL-Pgp was used as a control. Data are averaged from triplicate measurements with standard error bars.

Pgp structures with bound substrates

To capture different conformations during the transport cycle of Pgp, we mutated the catalytic glutamate residues in the Walker-B motifs of each NBD (E552Q/E1197Q) in Pgp335 and Pgp978. When Mg2+ATP is present, these mutations arrest the enzyme in an ATP-occluded, NBD-dimerized conformation that resembles the pre-hydrolysis state26. For Pgp971, we prepared the covalent complex in the presence of Mg2+ ATP and stabilized the resulting OF complex with vanadate in a post-hydrolysis intermediate state30,31 (Fig. 1c). CryoEM analysis of the respective covalent complexes in both OF and IF conformations provided eight high-resolution structures with zero, one, or two ligands bound. In addition, a control dataset with ATP, but without the substrate was collected for Pgp335 (Fig. 2, Extended Data Fig. 6 and 7, and Extended Data Tables 1 and 2).

High-resolution structures of Pgp with bound ligands.

For all mutants, IF and OF structures were obtained, with ligands revealed in the transmembrane region (grey shadow). For Pgp335 three different OF conformations were detected in one data set (OF335-nolig, OF335-1lig, OF335-2lig). To display the binding sites, the upper halves of TM 2,10,11 and 12 were removed from the densities. For the IF structures, TM9 was also removed for clarity. TM1 is shown in brown, with the covalently bound ligand in red and the non-crosslinked in green. Significant conformational changes were observed among these structures from which a further zoom-in of TM1 (black box region) was shown in Fig. 4 and Extended Data Fig. 8.

The IF-structure of Pgp335, solved at 3.8 Å resolution, shows clear densities for two substrate molecules: one crosslinked and one not (IF335-2lig, Fig. 2 and Extended Data Fig. 7 and 8). The non-crosslinked molecule is bound in a position that resembles the one previously reported in a crystal structure24. It interacts closely with the covalently-linked substrate, which is situated further up the membrane. Both ligand molecules are oriented along the vertical axis of the protein and positioned side-by-side through hydrophobic interactions. The uncleaved DNPT group of the non-crosslinked substrate is evident in the cryoEM density map. Two substrate molecules are also bound in the IF structure of Pgp978 (IF978-2lig); the non-covalently bound molecule is clearly visible at a similar position as in IF335-2lig, while the crosslinked substrate is located on the opposite side of the binding cavity near TM12 (Extended Data Fig. 8). Corroborating our labeling data (Fig. 1c), in the IF structure of Pgp971 (IF971-1lig) at 4.3 Å resolution, only the non-crosslinked substrate is bound at the approximate location seen for non-crosslinked substrate in IF335-2lig and IF978-2lig (Fig. 2, Extended Data Figs. 8 and 9e).

Multi-model cryoEM analyses of ATP-bound Pgp335 revealed three different OF conformations within a single dataset: ligand-free (40.4%), single-ligand-bound (33.3%), and double-ligand-bound (26.4%) (OF335-nolig, OF335-1lig and OF335-2lig), at 2.6 Å, 2.6 Å, and 3.1 Å resolution, respectively (Extended Data Fig. 10, Fig. 2). For Pgp978 and Pgp971, only OF conformations with a single bound ligand (OF978-1lig and OF971-1lig, Fig. 2) were detected at 2.9 Å and 3 Å resolution, respectively.

Substrate translocation pathway

In the OF structures with a single bound substrate, the central transmembrane helices TM1, TM6, TM7 and TM12 bulge outwards to accommodate the substrate in a small tunnel (Extended Data Fig. 11). Comparing the ligand positions between the IF and OF structures, in IF335-2lig to OF335-1lig, and IF978-2lig to OF978-1lig, the covalently attached ligand pivots around the respective Cys residue by almost 180°, moving it further up the translocation tunnel (Extended Data Figs. 8 and 9a,b, and Supplementary Videos 1 and 2). Our data do not reveal whether rotation of the ligand during the IF-OF conformational change is required for transport or if this reflects the fact that the ligand is covalently pinned to the crosslinking residue. However, the observation of ligand rotation illustrates how much space is available in the binding cavity during this conformational transition. The ligand position and orientation in OF335-1lig and OF978-1lig overall coincide well, with the latter shifting up slightly (Extended Data Fig. 9d). This finding indicates that the export tunnel is inherent to substrate translocation regardless of the initial binding position in the IF conformation. As such, our data suggest a single common pathway for substrate export in Pgp, contrary to the dual pseudosymmetric pathways suggested earlier28,32.

OF971-1lig captures a later-stage transport intermediate, in which the compound resides approximately 7 Å further up the translocation tunnel and is coordinated by highly conserved residues, including M74 and F78 (Figs. 2 and 3b, Extended Data Figs. 8 and 12). As no crosslinked substrate was detected in the corresponding IF structure, the substrate was likely transported to residue L971C before the crosslinking reaction could take place. Substrate entry from the extracellular side is less likely given the small opening of the substrate translocation pathway in all OF Pgp structures with or without bound substrate obtained here and previously26. In all substrate bound OF conformations methionines and phenylalanines at different positions seem to be important for ligand coordination in all of the Pgp structures (Extended Data Fig. 12). As expected, the substrate is held in the binding pocket mainly by the hydrophobic interactions throughout the different conformations (Extended Data Figs. 8 and 12). In the OF335-nolig structure, as well as in the virtually indistinguishable control, the translocation tunnel is collapsed, which shields the substrate binding pocket from the extracellular milieu and prevents re-entry from this side26.

Functional characterization of G72A mutant.

(a) TM1 breaks at the G72 position in the mid-transport states of Pgp. Shown is a close-up view of TM1 (brown) with side chain densities, the G72 position (green arrow), and the adjacently bound cyclic peptide AAC (red sticks) in the OF335-1lig structure. (b) Sequence alignment of TM1 in vertebrate Pgps including the highly conserved G72 residue (green arrow). (c) Level of expression of the G72A mutant relative to the WT- and CL-Pgp in S. cerevisiae cultures derived from the western blot analysis. (d) In vivo activity of the G72A mutant. The growth resistance of S. cerevisiae cultures transformed with either vector control plasmid, WT-Pgp, CL-Pgp, or the G72A mutant against several fungicides (doxorubicin, FK506, and valinomycin) relative to the YPD medium only (black bar) were analyzed. Compared to WT-Pgp and CL-Pgp, the G72A mutant severely compromised growth resistance against the tested drugs (p-values of <0.001 are indicated by ***). (e-g) ATPase activity of the purified G72A mutant. While basal ATP hydrolysis rates (in the absence or at low concentrations of drug) were indistinguishable between WT and G72A mutant proteins, stimulation of the ATPase activities by 1-3 µM FK506, 1-3 µM valinomycin, or 30-130 µM verapamil were severely impaired (p-values of <0.001 are indicated by ***).

Regulatory role of TM1

The series of IF and OF Pgp structures reveal a cascade of conformational changes in TM1, which displays significant plasticity throughout the different transport stages (Fig. 2, Extended Data Fig. 8, and Supplementary Videos 1, 2 and 3). In IF conformations, TM1 is a long, straight helix (Fig. 2 and Extended Data Fig. 8). However, in OF335-1lig, OF978-1lig, and OF971-1lig, TM1 swings out at A63, dilating the transmembrane passage and providing sufficient space for accommodating the bound ligand. At the same time, a pronounced kink at G72 on TM1 destabilizes the helix and the extracellular loop of Pgp between TM1 and TM2 and leads to a partial unwinding. This conformational deformation shields the intramembranous tunnel from the extracellular environment, with TM1 acting as a lid, almost parallel to the membrane (Fig. 2 and 4a, Extended Data Fig. 8). Strikingly, along with the upward movement of the ligand (from L335C to V978C to L971C), the TM1 loop is progressively lifted and restabilized. Ultimately, as shown previously26 and by our OF335-nolig structure, the helix is straightened upon the release of the ligand (Fig. 4a, Extended Data Fig. 8). Interestingly, the extracellular gate is sealed in all OF structures, which is especially surprising for OF971-1lig where the ligand is near the tunnel exit (Fig. Extended Data Fig. 11).

Tracing the substrate movement in Pgp.

Structures of OF Pgp from cryoEM (a) aligned with four snapshots from MD simulations (b) of substrate transport and release. TM1 is rendered in brown, with A63 and G72 in green, and M68, M74, and F78 in plum. The cyclic peptide is shown as red sticks. Significant conformational changes were observed for TM1 along with the upward movement and release of the substrate from OF Pgp.

Using OF978-1lig or OF335-1lig as initial configuration, we performed molecular dynamics (MD) simulations of Pgp embedded in a lipid bilayer to validate the proposed substrate escape pathway. In simulations starting from OF978-1lig at T = 400 K, we observed movement and release of the non-covalently bound substrate to the extracellular side moderated by interactions between the ligand and hydrophobic residues in the upper part of TM1 (Supplementary Video 3). The intermediate steps coincided well with the conformations observed in cryoEM (Fig. 4, Extended Data Figs. 13 and 14). Without any bias imposed, the substrate progressed to a position closely matching OF971-1lig (Fig. 2, 4b). Then, while TM1 gradually refolded from the intracellular side, the substrate wiggled out to form transient interactions with the still unstructured extracellular part of TM1 (Fig. 4). Importantly, the substrate escaped without Pgp experiencing a wide opening of the extracellular region. After the substrate left the exit tunnel, TM1 continued to refold to attain a straight helix configuration like OF335-nolig at the end of the simulation (Fig. 4). In a simulation starting from OF335-1lig performed at T = 330 K, a similar release pathway was observed (Extended Data Fig. 15). As the only difference, the substrate has not fully dissociated from the protein after 2.5 μs, with weak interactions persisting at the top part of TM1 from the extracellular side. Importantly, this is a configuration observed also in higher temperature simulations but with much shorter lifetime. Overall, with exception of TM1, motions in Pgp to accommodate the moving substrate were mostly limited to small dilations of the transmembrane helices, supporting our observations from cryoEM.

Our cryoEM structures and MD simulations revealed an important regulatory role of TM1 during substrate transport, emphasizing the helix-breaking G72. Therefore, we mutated this residue, which is highly conserved in mammalian Pgp (Fig. 3), to a helix-stabilizing alanine33 and tested the mutant’s ability to export structurally diverse fungicides and to convey drug resistance in vivo. Compared to the wild-type and CL-Pgp, the G72A mutant expressed normally in Saccharomyces cerevisiae (Fig. 3c), but showed significantly reduced growth resistance to the test drugs including doxorubicin, FK506, and valinomycin (Fig. 3d). As drug transport was severely impaired, we further investigated the mutant’s ATPase activity. In presence of transport substrates such as FK506, valinomycin, or verapamil the ATPase activity of WT Pgp was significantly stimulated; however, the G72A mutant displayed only low levels of basal ATPase activity (in the absence of drug) and the response to drug stimulation was greatly diminished over a range of concentrations tested (Fig. 3e-g). To reveal potential structural consequences responsible for the loss of function, we performed cryoEM on the Pgp335 G72A mutant after crosslinking the single cysteine L335C with AAC-DNTP and stabilizing the OF complex with Mg2+ ATP and vanadate in a post-hydrolysis intermediate state. After multi-model classification, two conformations emerged. The first conformation, resolved at 3.0 Å, shows no substrate bound in analogy to OF335-nolig. The second dominant conformation is, however, much more heterogenous, resulting in a reduced and anisotropic resolution of approximately 4.6 Å (Extended Data Fig. 16). Here, the NBDs are well resolved but several transmembrane helices are missing, including TM1 in which residue G72A is located. The structural disorder, observed in the transmembrane region, is likely a result of the tethered ligand, causing local distortion of the surrounding α-helices as TM1 may not be able to kink at G72 in the mutant, which could explain its impaired transport function. Collectively, the significant differences in drug resistance profiles and drug-stimulated ATP hydrolysis rates between the G72A mutant and WT-Pgp along with our structural findings substantiate and highlight the key role of TM1 for the regulation of substrate transport in Pgp.

Discussion

Our data suggest a mechanism for how Pgp mediates substrate transport. Transport commences in the IF conformation with binding of the substrate. Upon ATP-induced NBD-dimerization, the transporter converges into an OF conformation in which the intracellular gate is closed. The presence of the substrate in the OF conformation of Pgp dislocates TM1 at A63 and introduces a sharp kink at G72, forcing the helix to unwind partially between TM1 and TM2 (Fig. 4 and Extended Data Fig. 8). This enlarges the extracellular loop which shields the transmembrane tunnel from the extracellular environment and guides the substrate to move up the translocation tunnel, allowing TM1 to gradually restabilize. During the last stages of transport (OF971-1lig to OF335-nolig), the TM1-TM2 regulatory loop straightens, and the substrate-interacting hydrophobic residues (M74 and F78) rotate and pivot away from the tunnel, which facilitates the escape of the substrate from the transporter. Although OF971-1lig captured a post-hydrolysis state, we cannot rule out the possibility of ligand release prior to ATP hydrolysis, when it is not covalently bound. Notwithstanding, the proposed TM1 kinking and straightening mechanism of substrate expulsion would obviate the requirement for the Y-shaped, wide-open OF conformation reported for multiple bacterial ABC transporters34,35, suggesting a substrate-dependent transport mechanism in Pgp, or else a divergence of mechanisms among ABC transporters.

For the OF335-2lig structure, observed in the same data set as OF335-1lig and OF335-nolig, we hypothesize that the hydrophobic nature of the substrate promotes the binding of two substrate molecules to each other. MD simulations with this structure as a starting conformation reveal no separation of the two molecules, suggesting that they might be treated as one bigger substrate in the transport event. With the binding positions differing between OF335-1lig and OF335-2lig (Extended Data Fig. 9c, 12 and 16), neither the cryoEM structure nor the simulations allow conclusions about a possible transport mechanism of two ligands at this point.

How ABC transporters move substrates through the lipid bilayer has been a long-standing question36. The scarcity of high-resolution structures of substrate-bound OF conformations37, especially in the highly transient intermediate states, has severely limited our understanding of this vital process. Here, we were able to determine eight high-resolution structures of the archetypical ABC transporter Pgp, revealing the underlying molecular mechanics of substrate translocation. In general, the large number and diversity of ABC transporters, and the even wider variety of their substrates, make it unlikely that a universal substrate translocation mechanism will emerge. Our strategy of stalling the otherwise highly transient intermediates in this process through crosslinking and subsequent high-resolution cryoEM analysis could serve as a blueprint to understand the transport mechanisms of lipid transporters as well as other members of the transporter family.

Supporting information